---
title: Stuck with plastic
author: Janith Wanniarachchi
date: '2021-02-08'
slug: stuck-with-plastic
comments: no
cover: ''
imgs: []
justify: no
lastmod: '2021-06-05T12:02:10+05:30'
license: ''
readingTime: yes
single: no
description: Tidy Tuesday analysis of Plastic pollution
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
library(tidyverse)
library(ggfortify)
tuesdata <- tidytuesdayR::tt_load('2021-01-26')
knitr::opts_chunk$set(echo = TRUE)
```

# The background

## Introduction

**TLDR:** [*Sarah Sauve*](https://sarahasauve.wordpress.com) collected the data for the city of St. John's, to find out the companies and people that are contributing to the amount of plastic in the city. The data was collected during October 2020.*   

<!-- What we are going to do from this dataset today is to classify the companies and which area they can focus on to get into a company that is not harmful. -->



## Data Dictionary

The plastic is categorized by [recycling codes](https://en.wikipedia.org/wiki/Recycling_codes).

+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| variable       | class     | description                                                                                                                                                  |
+:===============+:==========+:=============================================================================================================================================================+
| country        | character | Country of cleanup                                                                                                                                           |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| year           | double    | Year (2019 or 2020)                                                                                                                                          |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| parent_company | character | Source of plastic                                                                                                                                            |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| empty          | double    | Category left empty count                                                                                                                                    |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| hdpe           | double    | High density polyethylene count (Plastic milk containers, plastic bags, bottle caps, trash cans, oil cans, plastic lumber, toolboxes, supplement containers) |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ldpe           | double    | Low density polyethylene count (Plastic bags, Ziploc bags, buckets, squeeze bottles, plastic tubes, chopping boards)                                         |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| o              | double    | Category marked other count                                                                                                                                  |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| pet            | double    | Polyester plastic count (Polyester fibers, soft drink bottles, food containers (also see plastic bottles)                                                    |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| pp             | double    | Polypropylene count (Flower pots, bumpers, car interior trim, industrial fibers, carry-out beverage cups, microwavable food containers, DVD keep cases)      |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ps             | double    | Polystyrene count (Toys, video cassettes, ashtrays, trunks, beverage/food coolers, beer cups, wine and champagne cups, carry-out food containers, Styrofoam) |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| pvc            | double    | PVC plastic count (Window frames, bottles for chemicals, flooring, plumbing pipes)                                                                           |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| grand_total    | double    | Grand total count (all types of plastic)                                                                                                                     |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| num_events     | double    | Number of counting events                                                                                                                                    |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| volunteers     | double    | Number of volunteers                                                                                                                                         |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+

# Data preparation

```{r}
skimr::skim_without_charts(tuesdata$plastic)
```

Several plastic counts columns contain NAs. Is this due to a mistake in entering data or was there actually no plastics of that category to be found? The most sound answer would be that there were no plastic counts to be found which is quite unlikely. We can recalculate the grand total with these imputed 0s.

Also let's focus on the year of 2019, since this study has finished on the October 2020. 

Interestingly, there are parent companies called Grand Total and Unbranded, and a country called EMPTY as well. 

```{r}
plastics <- tuesdata$plastics %>% 
  group_by(parent_company,country,year) %>% 
  mutate(across(empty:pvc,~replace_na(.x,0))) %>% 
  mutate(grand_total = sum(across(empty:pvc))) %>% 
  ungroup() %>% 
  filter(year == 2019) %>% 
  filter(country != "EMPTY" & 
           !(parent_company %in% c("Grand Total","Unbranded")) &
           grand_total != 0)
```


What if instead of this approach we decided to model the plastic counts using a Bayesian approach in order to impute the missing values?

For this article we will be using Stan as our Probabilistic Programming Backend. 