---
title: Stuck with plastic
author: Janith Wanniarachchi
date: '2021-02-08'
slug: stuck-with-plastic
categories:
  - Statistics
tags:
  - TidyTuesday
comments: no
cover: ''
imgs: []
justify: no
lastmod: '2021-06-05T12:02:10+05:30'
license: ''
readingTime: yes
single: no
Summary: Tidy Tuesday analysis of Plastic pollution
---


```{r setup, include=FALSE}
library(tidyverse)
tuesdata <- tidytuesdayR::tt_load('2021-01-26')
plastics <- tuesdata$plastics
knitr::opts_chunk$set(echo = TRUE)
```

# The background

## Introduction

**TLDR:** [*Sarah Sauve*](https://sarahasauve.wordpress.com)*. collected the data for the city of St. John's, to find out the companies and people that are contributing to the amount of plastic in the city. The data was collected during October 2020.*

From from the Tidy Tuesday Github page:

The data this week comes from [Break Free from Plastic](https://www.breakfreefromplastic.org) courtesy of [Sarah Sauve](https://sarahasauve.wordpress.com).

Sarah put together a nice [Blogpost](https://github.com/sarahsauve/TidyTuesdays/blob/master/BFFPDashboard/BlogPost.md) on her approach to this data, which includes cleaning the data and a Shiny app!

Per Sarah:

> I found out about [Break Free From Plastic](https://www.breakfreefromplastic.org/)'s Brand Audits through my involvement with the local [Social Justice Cooperative of Newfoundland and Labrador](https://www.sjcnl.ca/)'s Zero Waste Action Team.
>
> One of my colleagues and friends proposed an audit in St. John's, partially to contribute to the global audit and as part of a bigger project to understand the sources of plastic in our city. We completed our audit in October 2020 and are the first submission to BFFP from Newfoundland! You can find our data presented in [this](https://sarahsauve.shinyapps.io/BrandAuditDashboard/) Shiny dashboard.

> It's an interesting dataset, with lots of room to play around and so many options for visualization, plus plastic pollution is an important topic to talk about and raise awareness of! You can read BFFP's Brand Audit Reports for [2018](https://www.breakfreefromplastic.org/globalbrandauditreport2018/), [2019](https://www.breakfreefromplastic.org/globalbrandauditreport2019/) and [2020](https://www.breakfreefromplastic.org/globalbrandauditreport2020/) to get an idea of what they've done with the data.

## Data Dictionary

The plastic is categorized by [recycling codes](https://en.wikipedia.org/wiki/Recycling_codes).

+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| variable       | class     | description                                                                                                                                                  |
+:===============+:==========+:=============================================================================================================================================================+
| country        | character | Country of cleanup                                                                                                                                           |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| year           | double    | Year (2019 or 2020)                                                                                                                                          |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| parent_company | character | Source of plastic                                                                                                                                            |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| empty          | double    | Category left empty count                                                                                                                                    |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| hdpe           | double    | High density polyethylene count (Plastic milk containers, plastic bags, bottle caps, trash cans, oil cans, plastic lumber, toolboxes, supplement containers) |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ldpe           | double    | Low density polyethylene count (Plastic bags, Ziploc bags, buckets, squeeze bottles, plastic tubes, chopping boards)                                         |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| o              | double    | Category marked other count                                                                                                                                  |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| pet            | double    | Polyester plastic count (Polyester fibers, soft drink bottles, food containers (also see plastic bottles)                                                    |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| pp             | double    | Polypropylene count (Flower pots, bumpers, car interior trim, industrial fibers, carry-out beverage cups, microwavable food containers, DVD keep cases)      |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ps             | double    | Polystyrene count (Toys, video cassettes, ashtrays, trunks, beverage/food coolers, beer cups, wine and champagne cups, carry-out food containers, Styrofoam) |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| pvc            | double    | PVC plastic count (Window frames, bottles for chemicals, flooring, plumbing pipes)                                                                           |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| grand_total    | double    | Grand total count (all types of plastic)                                                                                                                     |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| num_events     | double    | Number of counting events                                                                                                                                    |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+
| volunteers     | double    | Number of volunteers                                                                                                                                         |
+----------------+-----------+--------------------------------------------------------------------------------------------------------------------------------------------------------------+

# At a glance

## Skim, Skim. Skim

```{r}
skimr::skim(plastics)
```

What are the ones that have grand total missing
```{r}
plastics %>% filter(is.na(grand_total)) %>% knitr::kable()
```

So Sri Lanka havent really given a lot of data for this, even though there has been.. 1 counting event. Wait how many rows are there with the num_events set to 1 or even worse 0

```{r}
plastics %>% filter(num_events == 1) %>% skimr::skim()
```

What if we replaced the NAs of the empty:pvc to zero and then recalculated the grand_totals

```{r}
plastics_ <- plastics %>% 
  mutate(across(empty:pvc,~replace_na(.x,0))) %>% 
  mutate(grand_total = sum(across(empty:pvc)))

```


So that's done. What about volunteers. 

```{r}
plastics_ %>% 
  filter(is.na(volunteers)) %>% 
  group_by(country,year) %>% 
  count()
```

What the hell they are all from Korea?
Why? And it's just the year of 2020

```{r}
plastics_ %>% 
  filter(is.na(volunteers)) %>% 
  filter(across(empty:pvc, ~is.na(.x)))
```

So volunteers arent an issue at all. Let's leave them at it. 

Now let's focus on the year and country distribution

```{r}
plastics_ %>% 
  count(country,year,sort=TRUE)
```

Wait what, there's an EMPTY country section? 

```{r}
plastics_ %>% 
  filter(country == "EMPTY")
```

Oh great, there's a parent_company called Unbranded and Grand Total

```{r}
plastics_ %>% 
  filter(grand_total == 0)
```

Well crap there are useful information in the EMPTY country as well. 